Thesis Title

### 2022-10-31 chapters/chapter_2.tex
Updates from Overleaf
Add abstract - first draft; Minor copyediting

In this publication \ac{SMg} was applied to nine body fluid samples and one tissue sample from patients at the University Medical Center Groningen (UMCG) with varying degrees of contamination: one sample from peritoneal fluid, five from pus , two from synovial fluid of knees with prosthesis, one from sputum and one from a bone biopsy. The results of microbial identification through whole genome sequencing (\ac{WGS}) and \ac{SMg} were compared to standard culture-based microbiological methods. 
In order to evaluate and compare the accuracy and reliability of the bioinformatics analyses in providing the closest results to culture and \ac{WGS} of any cultured isolates, three different bioinformatic pipelines (two commercially and one freely available) were used. Most pathogens identified by culture were also identified through metagenomics, but substantial differences were noted between the taxonomic classification tools.

My contribution to this publication included the bioinformatics analysis of all the samples using a unix-based approach. I performed quality assessment and quality control of the \ac{WGS} and \ac{SMg} data, the removal of host sequencing from the samples, and the taxonomic identification of the remaining reads in each sample though 3 different methods: MetaPhlan2, Kraken and MIDAS. Gene detection directly from the reads for bacterial typing was also performed using metaMLST, ReMatCh, and Bowtie2 and Samtools. Finally, the reads were assembled using the SPAdes genome assembler, with and without metagenomic mode according to the sample being processed.

>>>>

In this publication, \ac{SMg} was applied to nine body fluid samples and one tissue sample from patients at the University Medical Center Groningen with varying degrees of contamination: one sample from peritoneal fluid, five from pus, two from synovial fluid of knees with a prosthesis, one from sputum and one from a bone biopsy. The results of microbial identification through whole genome sequencing (\ac{WGS}) and \ac{SMg} were compared to standard culture-based microbiological methods. 
In order to evaluate and compare the accuracy and reliability of the bioinformatics analyses in providing the closest results to culture and \ac{WGS} of any cultured isolates, three different bioinformatic pipelines (two commercially and one freely available) were used. Most pathogens identified by culture were also identified through metagenomics, but substantial differences were noted between the taxonomic classification tools.

My contribution to this publication included the bioinformatics analysis of all the samples using a Unix-based approach. I performed quality assessment and quality control of the \ac{WGS} and \ac{SMg} data, the removal of host sequencing from the samples, and the taxonomic identification of the remaining reads in each sample through 3 different methods: MetaPhlan2, Kraken and MIDAS. Gene detection directly from the reads for bacterial typing was also performed using metaMLST, ReMatCh, Bowtie2 and Samtools. Finally, the reads were assembled using the SPAdes genome assembler, with and without metagenomic mode according to the sample being processed.

### 2022-10-31 chapters/chapter_3.tex
Add abstract - first draft; Minor copyediting

As referenced in section \ref{ssec:survaillance}, sequencing has become a common tool in surveillance and infection prevention, when combined with epidemiological data, have undoubtedly provided immeasurable insights regarding identification of potential sources of pathogenicity and transmission pathways. Shotgun metagenomic (SMg) approaches, just like in a clinical setting,  have been a growing interest to deliver relevant results without a priori knowledge of what to expect from a particular environmental sample.

In this publication, second (see \ref{ssec:2nd_gen_seq}) and third (see \ref{ssec:3rd_gen_seq}) generation sequencing SMg has been applied to eight concentrated water samples collected the University Medical Center Groningen. In one of the samples, the novel detection of an \textit{mcr-5} gene, named \textit{mcr-5.4}, is reported. To the best of our knowledge, this is the first time that this gene, a mobile colistin resistance (\textit{mcr}) determinant, has been recovered from a hospital water environment, with analysis suggesting the order of \textit{Pseudomonadales} as the most probable host.

My contribution to this publication included the bioinformatics analysis of the \textit{mcr-5.4} carrying sample thorough hybrid assembly using metaSPAdes. The resulting assembled contings were binned with the MaxBin2 tool and the bin having the sequence carrying the gene of interest was taxonomically characterised with Kraken2.

>>>>

As referenced in Chapter \ref{ch:introduction}, section \ref{ssec:_intro_survaillance}, sequencing has become a common tool in surveillance and infection prevention, when combined with epidemiological data, have undoubtedly provided immeasurable insights regarding the identification of potential sources of pathogenicity and transmission pathways. \ac{SMg} approaches, just like in a clinical setting, have been a growing interest to deliver relevant results without a priori knowledge of what to expect from a particular environmental sample.

In this publication, second (see Chapter \ref{ch:introduction}, section \ref{ssec:_intro_2nd_gen_seq}) and third (see Chapter \ref{ch:introduction}, section \ref{ssec:_intro_3rd_gen_seq}) generation sequencing \ac{SMg} has been applied to eight concentrated water samples collected the University Medical Center Groningen. In one of the samples, the novel detection of an \textit{mcr-5} gene, named \textit{mcr-5.4}, is reported. To the best of our knowledge, this is the first time that this gene, a mobile colistin resistance (\textit{mcr}) determinant, has been recovered from a hospital water environment, with analysis suggesting the order of \textit{Pseudomonadales} as the most probable host.

My contribution to this publication included the bioinformatics analysis of the \textit{mcr-5.4} carrying sample thorough hybrid assembly using metaSPAdes. The resulting assembled contigs were binned with the MaxBin2 tool and the bin having the sequence carrying the gene of interest was taxonomically characterised with Kraken2.

### 2022-10-31 chapters/chapter_4.tex
Add abstract - first draft; Minor copyediting

\ac{DENV} represents a public health threat and economic burden in affected countries. The risk of exposure to \ac{DENV} is increasing, not only because of travel to endemic regions, but also due to the broader dissemination of the mosquito vector, making the burden of dengue very significant.

The availability of genomic data is key to understanding viral evolution and dynamics, supporting improved control strategies. Currently, the use of second-generation sequencing technologies, which can be applied both directly to patient samples (shotgun metagenomics) and to PCR-amplified viral sequences (amplicon sequencing), is the most informative approach to monitor viral dissemination and genetic diversity by providing, in a single methodological step, identification and characterization of the whole viral genome at the nucleotide level. This makes \ac{DENV} identification and characterization through genomic analysis by developing a software where the lessons learned in Chapters \ref{ch:paper1} and \ref{ch:paper2} are applied.

@@

My contribution to this publication included the design, implementation and optimisation of the DEN-IM the workflow, including the creation of  the Docker containers for all dependencies. Two databases, one comprising 3830 \ac{DENV} sequences for the retrieval of the reads of interest from the input samples, and a second comprising of 161 sequences representing the genetic diversity of all \ac{DENV} sero and genotypes were constructed by me.  Additionally, I've also wrote the manuscript.

>>>>

\ac{DENV} represents a public health threat and economic burden in affected countries. The risk of exposure to \ac{DENV} is increasing, not only because of travel to endemic regions but also due to the broader dissemination of the mosquito vector, making the burden of dengue very significant.

The availability of genomic data is key to understanding viral evolution and dynamics and supporting improved control strategies. Currently, the use of second-generation sequencing technologies, which can be applied both directly to patient samples (shotgun metagenomics) and to PCR-amplified viral sequences (amplicon sequencing), is the most informative approach to monitor viral dissemination and genetic diversity by providing, in a single methodological step, identification and characterization of the whole viral genome at the nucleotide level. This makes \ac{DENV} identification and characterization through genomic analysis by developing software where the lessons learned in Chapters \ref{ch:paper1} and \ref{ch:paper2} are applied.

@@

My contribution to this publication included the design, implementation and optimisation of the DEN-IM workflow, including the creation of  the Docker containers for all dependencies. Two databases, one comprising 3830 \ac{DENV} sequences for the retrieval of the reads of interest from the input samples, and a second comprising 161 sequences representing the genetic diversity of all \ac{DENV} sero and genotypes were constructed by me.  Additionally, I've also written the manuscript.

### 2022-10-31 chapters/chapter_5.tex
Add abstract - first draft; Minor copyediting

This chapter is a reproduction of the following submitted manuscript for publication in GigaScience:

@@

To test assembler performance with an even more complex dataset, we used the 12 strain BMock12 community standard. This sample includes a non-even distribution of species and several closely related sets: two replicons of Halomonas sp. (ANIb=0.98), three replicons of the Micromonospora genus (average ANIb=0.85) and two replicons of Marinobacter sp (ANIb=0.78). Furthermore, and to represent a mock community trying to reproduce an existing microbiome, the NIBSC Gut DNA Reference Gut-Mix-RR and Gut-Mix-HiLo community standard were analysed, consisting of 20 common gut microbiome strains in an even and staggered composition respectively.

>>>>

This chapter is a reproduction of the following manuscript accepted for publication in GigaScience:

@@

To test assembler performance with an even more complex dataset, we used the 12-strain BMock12 community standard. This sample includes a non-even distribution of species and several closely related sets: two replicons of Halomonas sp. (ANIb=0.98), three replicons of the Micromonospora genus (average ANIb=0.85) and two replicons of Marinobacter sp (ANIb=0.78). Furthermore, and to represent a mock community trying to reproduce an existing microbiome, the NIBSC Gut DNA Reference Gut-Mix-RR and Gut-Mix-HiLo community standards were analysed, consisting of 20 common gut microbiome strains in an even and staggered composition respectively.

### 2022-10-31 chapters/chapter_6.tex
Add abstract - first draft; Minor copyediting

This chapter is a reproduction of the following submitted manuscript for publication in eLife:

C. I. Mendes, F. Maguire, A. Manuele, D. Fornika, S. Tausch, T. Le-Viet, J. Phelan, C. Meehan, A. Raphenya, B. Alcock, E. Culp, A. McArthur, M. Feldgarden, G. Tyson, M. Galas, J. Campos, A. Witney, D. Aanensen, A. Black, L. Katz, P. Oluniyi, I. Olawoye, R. Timme, A. T. Vasconcelos, A. Page, D. MacCannell, E. Griffiths. 
hAMRonization: Enhancing antimicrobial resistance prediction using PHA4GE standards and specification

As discussed in Chapters \ref{ch:introduction} \ref{ch:paper1} and \ref{ch:paper2}, \ac{SMg} can offer relatively unbiased pathogen detection and characterisation, potentially able to provide antimicrobial resistance and virulence profiling in a single methodological step. The rise of \ac{AMR} poses a major threat to human health around the world.

@@

In this chapter we present a a standardised output specification for the reporting of genes or variants potentially conferring \ac{AMR}. This specification aims to standardise the reporting of these results, regardless of the tool and/or database used. To ease its utilisation, we've packaged this specification into hAMRonization, a command-line utility that is able to aggregate results form a wide variety of \ac{AMR} detection tools, both species-agnostic and species-specific, providing a unified report tabular form, JSON or through an interactive HTML file that can be opened within the browser for navigable data exploration.

We aim that with hAMRonization, stakeholders, such as clinical practitioners, will be more efficiently informed of potential \ac{AMR} genes of variants in a given dataset containing genomic information. Ultimately, the possibility of using multiple \ac{AMR} detection tools, with multiple \ac{AMR} databases, increases evidence, which increases confidence in the results obtained. Likewise, differences in databases can be evidenced fir better interpretability.

My contribution to this publication included the development of the antimicrobial detection contextual data specification package, including it’s conversion and availability in a machine-applicable JSON format. It also includes the design and implementation of the hAMRonization package, results analysis, manuscript production and editing.

>>>>

This chapter is a reproduction of the following manuscript to be submitted for publication in eLife:

C. I. Mendes, F. Maguire, A. Manuele, D. Fornika, S. Tausch, T. Le-Viet, J. Phelan, C. Meehan, A. Raphenya, B. Alcock, E. Culp, A. McArthur, M. Feldgarden, G. Tyson, M. Galas, J. Campos, A. Witney, D. Aanensen, A. Black, L. Katz, P. Oluniyi, I. Olawoye, R. Timme, A. T. Vasconcelos, A. Page, D. MacCannell, E. Griffiths. 
hAMRonization: Enhancing antimicrobial resistance prediction using PHA4GE standards and specification.

As discussed in Chapters \ref{ch:introduction} \ref{ch:paper1} and \ref{ch:paper2}, \ac{SMg} can offer relatively unbiased pathogen detection and characterisation, potentially able to provide antimicrobial resistance and virulence profiling in a single methodological step. The rise of \ac{AMR} poses a major threat to human health worldwide.

@@

In this chapter, we present a standardised output specification for the reporting of genes or variants potentially conferring \ac{AMR}. This specification aims to standardise the reporting of these results, regardless of the tool and/or database used. To ease its utilisation, we've packaged this specification into hAMRonization, a command-line utility that is able to aggregate results from a wide variety of \ac{AMR} detection tools, both species-agnostic and species-specific, providing a unified report tabular form, JSON or through an interactive HTML file that can be opened within the browser for navigable data exploration.

We aim that with hAMRonization, stakeholders, such as clinical practitioners, will be more efficiently informed of potential \ac{AMR} genes of variants in a given dataset containing genomic information. Ultimately, the possibility of using multiple \ac{AMR} detection tools, with multiple \ac{AMR} databases, increases evidence, which increases confidence in the results obtained. Likewise, differences in databases can be evidenced for better interpretability.

My contribution to this publication included the development of the antimicrobial detection contextual data specification package, including its conversion and availability in a machine-applicable JSON format. It also includes the design and implementation of the hAMRonization package, results analysis, manuscript production and editing.

### 2022-10-31 chapters/chapter_7.tex
Add abstract - first draft; Minor copyediting

As mentioned in Chapter \ref{ch:introduction}, the \ac{SARS-CoV-2} has brought a new meaning to genomic surveillance, with currently, over 8 million complete viral sequences are available at GISAID\footnote{\url{https://www.gisaid.org/}}, being one of the most highly sequenced genomes of any organism on the planet. 
This richness in genomic information has been basal to identifying new variants of risk and new variants of concern with a myriad of different origins, identifying routes of transmission across borders, including the identification of "super-spreaders" events, and informing infection control measures.

@@

My contribution to this publication included the development of the SARS-CoV-2 contextual data specification package, including it's conversion and availability in a machine-applicable JSON format. I've also maintain the public repository where the data specification package is hosted\footnote{\url{https://github.com/pha4ge/SARS-CoV-2-Contextual-Data-Specification/}} Additionally, I've also contributed to the manuscript production and editing.

>>>>

As mentioned in Chapter \ref{ch:introduction}, the \ac{SARS-CoV-2} has brought a new meaning to genomic surveillance, with currently, over 8 million complete viral sequences available at GISAID\footnote{\url{https://www.gisaid.org/}}, being one of the most highly sequenced genomes of any organism on the planet. 
This richness in genomic information has been basal to identifying new variants of risk and new variants of concern with a myriad of different origins, identifying routes of transmission across borders, including the identification of "super-spreaders" events, and informing infection control measures.

@@

My contribution to this publication included the development of the SARS-CoV-2 contextual data specification package, including its conversion and availability in a machine-applicable JSON format. I've also maintained the public repository where the data specification package is hosted\footnote{\url{https://github.com/pha4ge/SARS-CoV-2-Contextual-Data-Specification/}} Additionally, I've also contributed to the manuscript production and editing.

### 2022-10-31 chapters/chapter_8.tex
Add abstract - first draft; Minor copyediting

We propose collaborative software testing as an opportunity to continuously engage software users, developers, and students to unify scientific work across domains. As automated software testing remains underused in scientific software, our set of recommendations not only ensures appropriate effort can be invested into producing high quality and robust software but also increases engagement in its sustainability.

@@

Computational algorithms have become an essential component of research, with great efforts by the scientific community to raise standards on development and distribution of code. Despite these efforts, sustainability and reproducibility are major issues since continued validation through software testing is still not a widely adopted practice. Here, we report seven recommendations that help researchers implement software testing in microbial bioinformatics. We have developed these recommendations based on our experience from a collaborative hackathon organised prior to the American Society for Microbiology Next Generation Sequencing (ASM NGS) 2020 conference. We also present a repository hosting examples and guidelines for testing, available from \url{https://github.com/microbinfie-hackathon2020/CSIS}.

@@

In the field of microbial bioinformatics, good software engineering practises are not yet widely adopted. Many microbial bioinformaticians start out as (micro)biologists and subsequently learn how to code. Without abundant formal training, a lot of education about good software engineering practices comes down to an exchange of information within the microbial bioinformatics community. This paper serves as a resource that could help microbial bioinformaticians get started with software testing if they have not had formal training.

>>>>

We propose collaborative software testing as an opportunity to continuously engage software users, developers, and students to unify scientific work across domains. As automated software testing remains underused in scientific software, our set of recommendations not only ensures appropriate effort can be invested into producing high-quality and robust software but also increases engagement in its sustainability.

@@

Computational algorithms have become an essential component of research, with great efforts by the scientific community to raise standards on the development and distribution of code. Despite these efforts, sustainability and reproducibility are major issues since continued validation through software testing is still not a widely adopted practice. Here, we report seven recommendations that help researchers implement software testing in microbial bioinformatics. We have developed these recommendations based on our experience from a collaborative hackathon organised prior to the American Society for Microbiology Next Generation Sequencing (ASM NGS) 2020 conference. We also present a repository hosting examples and guidelines for testing, available from \url{https://github.com/microbinfie-hackathon2020/CSIS}.

@@

In the field of microbial bioinformatics, good software engineering practices are not yet widely adopted. Many microbial bioinformaticians start out as (micro)biologists and subsequently learn how to code. Without abundant formal training, a lot of education about good software engineering practices comes down to an exchange of information within the microbial bioinformatics community. This paper serves as a resource that could help microbial bioinformaticians get started with software testing if they have not had formal training.
